---
title: "Lab 02: Simple Regression"
author: "PSYC 7804"
date: "Spring 2021"
output: 
  beamer_presentation:
    theme: "Madrid"
    colortheme: "beaver"
editor_options: 
  chunk_output_type: console
---
  
```{r include = FALSE}
library(knitr)
opts_chunk$set(class.output='sh', comment = "")
```

## Packges for Today (Install if Needed)

```{r, message = FALSE}
library(knitr)
library(tidyverse)
library(haven)
library(ggplot2)
library(psych)
```

## Overview

Functions: 
  
* `cor()` and `cov()`: compute correlation/correlation with arguments:
  + `use`: how to treat missing data, specify "everything", "complete.obs", "pairwise.complete.obs"
  + `method`: which type of correlation, specify "pearson", "kendall", or "spearman"

* `geom_smooth()`: adding best-fit lines to a scatterplot with `ggplot`, with arguments:
  + `method`: `lm` for a linear model or `loess`
  + `formula`: specify the form of a linear model (e.g., quadratic, cubic)

* `lm()`: fit a linear regression model

* `cor.test()`: hypothesis test for the correlation coefficient

## Objectives

* Understand simple linear regression
* Calculate regression parameters
* Understand how we assess the fit of a regression model
  + Total Sum of Squares
  + Model Sum of Squares
  + Residual Sum of Squares
  + F test
  + $R^2$
* Know how to interpret the outputs
* Interpret a regression mode

## Linear Regression

* Regression analysis is used to predict the value of one variable (the dependent variable) based on other variables (the independent variables).
  + Dependent variable: denoted $Y$
  + Independent variables: denoted $X_1,X_2,\ldots,X_K$
* It is a hypothetical model of the linear relationship between the DV and a set of IVs.
* **Simple** linear regression has one IV.
* **Multiple** linear regression has multiple IVs.

## Sales example

A record company boss was interested in predicting record sales from advertising.

```{r, eval = TRUE}
dat <- read_sav("dat_sales.sav")
dat %>% glimpse
```

* Data: 200 different album releases
* `sales`: Sales (CDs and Downloads) in the week after release
* `adverts`: The amount (in £s) spent promoting the record before release.

## gg-Plotting with Best-Fit Lines
\small
Today, we'll look at the relationship between `adverts` and `sales`. 

Let's first construct a scatterplot to of `adversts` and `sales`:

```{r, fig.height = 4.5}
ggplot(dat, aes(adverts, sales)) + geom_point()
```

## gg-Plotting with Best-Fit Lines

We can add different types of best-fit lines using `geom_smooth()`. 

The `method` option tells `ggplot` which type of best-fit line to draw. The `lm` option indicates a linear model (simple linear regression by default).

* In `geom_smooth`, "formula" is not necessary, but prevents a message

\footnotesize
```{r, fig.height = 3.5}
ggplot(dat, aes(adverts, sales)) + geom_point() + 
  geom_smooth(method = lm, formula = y ~ x) 
```

## gg-Plotting with Best-Fit Lines

By default, `ggplot` adds an error band around the best-fit line. We can turn this off using the argument `se = FALSE`.

```{r, fig.height = 4.5}
ggplot(dat, aes(adverts, sales)) + geom_point() + 
  geom_smooth(method = lm, formula = y ~ x, se = FALSE)
```

## Correlations

The `cov()` and `cor()` functions are the easiest ways to calculate the covariance and the correlation.

```{r}
cov(dat$adverts, dat$sales)
cor(dat$adverts, dat$sales,use = "complete.obs")

```

We don't have any missing data in this example, but it's common to run into missingness when computing correlations. The `use` argument tells `R` how to handle missing data. For example, `use = "complete.obs"` will compute the correlation based only on the complete cases.  

## Correlations: Hypothesis Test

To do a hypothesis test of the Pearson product-moment (PPM) correlation, use the `cor.test()` function:
  
```{r}
cor.test(dat$adverts, dat$sales)
```


## Correlations: Hypothesis Test

Not surprisingly, this correlation is significantly different from 0, $r = .58, t(198) = 9.98, p < .001 (CI = [.48, .66])$. This function also gives us the confidence interval.

## Regression

Finally, let's run a linear regression with `sales` as the DV and `adverts` as the IV. This will give us the equation for the linear best-fit line that we drew on the scatterplot earlier.

Linear regression in `R` is done with the `lm` function. I like to take a `summary()` of this output right away because it gives us the summary coefficient table and information about $R^2$, etc. `lm` takes as input a formula of the form: "DV ~ IV", and you can specify the dataset as the second argument.

```{r, eval = FALSE}
summary(lm(sales ~ adverts, dat))
```

Output on next slide.

## Regression

\footnotesize
```{r, echo = FALSE}
res <- lm(sales ~ adverts, dat)
summary(res)
```

## Regression

$$y = \beta_0 + \beta_1\times x +\epsilon$$

* Variables
  + X = Independent Variable
  + Y = Dependent Variable

* Parameters: 
  + $\beta_0$ = Y-Intercept
  + $\beta_1$ = Slope

* $\epsilon$ = residual error
  + Residuals are assumed to be normally distributed

## Understanding Regression Coefficients

The output tells us that $\widehat{sales} = 134.14 + 0.096\times adverts$. 

* For `adverts` = 0, the predicted value of `sales` would be 134.14. 

* For a one-unit increase in `adverts`, `sales` are expected to increase by 0.096 on average.


These interpretations assume that a straight line is the best representation of the relationship between `adverts` and `sales`. 

## Test the relationship between IV and DV

* The t-test results of $\beta_1$ (slope coefficient) suggested that is significant relationship between these 2 variables, $\beta_1=0.096$, $t(198) = 9.98$, $p<.001$. 
  + The t-test is conducted to test the hypothesis below:
  + Null hypothesis: $\beta_1=0$.
  + Alternative hypothesis: $\beta_1\ne0$

* To find the 95% CI of $\beta_0$ and $\beta_1$, we can use the `confint` function

\small
```{r}
res <- lm(sales ~ adverts, dat)
confint(res,level = 0.95)
```

## How Good is the Model?

* The regression line is only a model based on the data, that might not reflect reality, we need some way of testing how well the model fits the observed data.

* SST: Total variability (variability between scores and the mean)
* SSE: Residual/Error variability (variability between the regression model and the actual data)
* SSR: Model variability (difference in variability between the model and the mean)

**If the model results in better prediction than using the mean, then we expect SSR to be much greater than SSE**

## Testing the Model
\footnotesize
1. ANOVA

* $F = \frac{MSR}{MSE}$
```{r}
anova(res)
```

2. $R^2$: The proportion of variance accounted for by the regression model

* $R^2 = \frac{SSR}{SST}$

* Adjusted $R^2$: an adjustment that penalizes the addition of extraneous predictors to the model.

##  Reporting results

Simple regression analysis shows that advertising budget significantly predicted higher record sales ($b = 0.10, p < .001$). The results of the regression indicated the predictor explained 33.5% of the variance ($R^2=.335$,$F(1, 198) = 99.59$, $p<.001$).
